# `gwzinbr()`: Model Fitting {#sec-gwzinbr}

## Overview

`gwzinbr()` is the main modelling function. Given a dataset with spatial coordinates
and a bandwidth $h$ (from `Golden()`), it fits a geographically weighted regression
model for count data and returns a rich list of local estimates, standard errors,
significance tests, and goodness-of-fit measures.

## Function signature

```r
gwzinbr(
  data,
  formula,
  xvarinf    = NULL,
  weight     = NULL,
  lat,
  long,
  grid       = NULL,
  method,
  model      = "zinb",
  offset     = NULL,
  distancekm = FALSE,
  force      = FALSE,
  int_inf    = TRUE,
  maxg       = 100,
  h          = NULL
)
```

## Parameter reference

### Data and model specification

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `data` | `data.frame` | — | Dataset (plain data frame, not `sf`). |
| `formula` | `formula` | — | Formula for the count component. Example: `cases ~ x1 + x2`. |
| `xvarinf` | `character` | `NULL` | Column name(s) for zero-inflation covariates. `NULL` = no zero-inflation component. |
| `weight` | `character` | `NULL` | Sample weight column. `NULL` = equal weights. |
| `offset` | `character` | `NULL` | Log-scale offset column. `NULL` = zero offset. |
| `model` | `character` | `"zinb"` | `"zinb"`, `"zip"`, `"negbin"`, or `"poisson"`. |
| `force` | `logical` | `FALSE` | Force the chosen model even when zero-inflation is not needed. |
| `int_inf` | `logical` | `TRUE` | Include an intercept in the zero-inflation component. |
| `maxg` | `integer` | `100` | Max iterations for the zero-inflation IRLS subproblem. |

### Spatial specification

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `lat` | `character` | — | Latitude / Y-coordinate column name. |
| `long` | `character` | — | Longitude / X-coordinate column name. |
| `distancekm` | `logical` | `FALSE` | Convert distances to km (multiply by 111). Use with decimal degrees only. |

### Bandwidth and prediction grid

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `h` | `numeric` | `NULL` | Bandwidth value. Use the output of `Golden()$min_bandwidth`. If `NULL`, fitting proceeds with all observations equally weighted (global model). |
| `grid` | `data.frame` | `NULL` | Additional locations for prediction. Must contain the same `lat`/`long` columns as `data`. Predictions are stored in `$param_estimates_grid`. |

## Supported model combinations

| `model` | `xvarinf` | `force` | Result |
|---------|-----------|---------|--------|
| `"zinb"` | `NULL` | `FALSE` | GWNBR (falls back automatically if no excess zeros) |
| `"zinb"` | `"var"` | `TRUE` | GWZINBR — full zero-inflated NB |
| `"negbin"` | any | — | GWNBR (zero inflation ignored) |
| `"zip"` | `NULL` | `FALSE` | GWZIP (falls back to GWPoisson if no excess zeros) |
| `"poisson"` | any | — | GWR-Poisson |

::: {.callout-note}
## Automatic model fallback
When `force = FALSE` and the expected number of zeros under the fitted NB
distribution meets or exceeds the observed zeros, the package prints a note
and silently drops the zero-inflation component. Set `force = TRUE` to
override this behaviour.
:::

## Return value {#sec-output-overview}

`gwzinbr()` returns a named list with **22 elements**. They are detailed in full
in @sec-output. Here is a quick-reference table:

| Element | Type | Description |
|---------|------|-------------|
| `$bandwidth` | scalar | Bandwidth value used |
| `$measures` | named vector | AIC, AICc, deviance, pseudo-R², adj. pseudo-R² |
| `$gwr_param_estimates` | matrix $N \times (p+1)$ | Local coefficient estimates |
| `$alpha_estimates` | matrix $N \times 2$ | Local $\alpha$ (overdispersion) estimates |
| `$residuals` | matrix $N \times 5$ | Residuals (multiple types) |
| `$t_test_gwr_param_estimates` | matrix | $t$-statistics and $p$-values per location |
| `$non_stationary_test` | data frame | Non-stationarity test per covariate |
| `$global_param_estimates` | matrix | Global model coefficients (reference) |
| `$qntls_gwr_param_estimates` | matrix | Quantiles of local estimates |
| `$descript_stats_gwr_param_estimates` | matrix | Mean, SD, min, max of local estimates |
| `$qntls_gwr_se` | matrix | Quantiles of local standard errors |
| `$descript_stats_gwr_se` | matrix | Descriptive stats of local SE |
| `$variance_covariance_matrix` | matrix | Variance-covariance of global estimates |
| `$qntls_gwr_zero_infl_param_estimates` | matrix | Quantiles of zero-inflation estimates |
| `$descript_stats_gwr_zero_infl_param_estimates` | matrix | Descriptive stats of ZI estimates |
| `$t_test_gwr_zero_infl_param_estimates` | matrix | $t$-tests for ZI component |
| `$qntls_gwr_zero_infl_se` | matrix | Quantiles of ZI standard errors |
| `$descript_stats_gwr_zero_infl_se` | matrix | Descriptive stats of ZI SE |
| `$non_stationary_test_zero_infl` | data frame | Non-stationarity test for ZI component |
| `$analysis_max_like_zero_infl_param_estimated` | matrix | MLE of ZI parameters |
| `$analysis_max_like_gof_measures` | matrix | GoF for ZI global model |
| `$param_estimates_grid` | matrix | Predictions at `grid` locations |

## Minimal example

```{r}
#| label: gwzinbr-minimal
#| eval: false

library(gwzinbr)
data(southkorea_covid19)

mod <- gwzinbr(
  data       = southkorea_covid19,
  formula    = n_covid1 ~ diff_sd,
  lat        = "y", long = "x",
  offset     = "ln_total",
  model      = "negbin",
  method     = "adaptive_bsq",
  h          = 230,
  distancekm = TRUE,
  force      = TRUE
)

mod$bandwidth
mod$measures
```

## Full example (all arguments)

```{r}
#| label: gwzinbr-full
#| eval: false

mod_full <- gwzinbr(
  data       = southkorea_covid19,
  formula    = n_covid1 ~ Morbidity + high_sch_p + Healthcare_access +
                          diff_sd + Crowding + Migration + Health_behavior,
  xvarinf    = NULL,       # no zero-inflation component
  weight     = NULL,       # equal weights
  lat        = "y",
  long       = "x",
  grid       = NULL,       # no prediction grid
  method     = "adaptive_bsq",
  model      = "negbin",
  offset     = "ln_total",
  distancekm = TRUE,
  force      = TRUE,
  int_inf    = TRUE,
  maxg       = 100,
  h          = 230         # from Golden()
)
```

## Notes on runtime

`gwzinbr()` fits a separate weighted GLM for each of the $N$ spatial units. For
large datasets or complex models (ZINB with many covariates), runtime can be
substantial. As a rough guide:

| $N$ | Model | Approximate time |
|-----|-------|-----------------|
| < 100 | NB | < 1 min |
| 250 | NB | 2–10 min |
| 250 | ZINB | 10–30 min |
| > 500 | NB | 30+ min |

Cache the result with `saveRDS()` immediately after fitting:

```r
saveRDS(mod, "cache/model_gwzinbr.rds")
mod <- readRDS("cache/model_gwzinbr.rds")
```
