# Diagnostics {#sec-diagnosticos}

After fitting a `gwzinbr()` model, four types of diagnostic questions arise:

1. **Is the model well-calibrated?** — goodness-of-fit measures
2. **Are the residuals spatially random?** — Moran I test
3. **Does spatial variation matter?** — non-stationarity tests
4. **Was zero-inflation necessary?** — zero excess check

## Goodness-of-fit

```{r}
#| label: gof-demo
#| eval: false

mod$measures
#   deviance       AIC      AICc    pct_ll adj_pct_ll
#     94.83    745.26    756.76      0.93       0.90
```

Compare the GWR model against the global Negative Binomial:

| Metric | Global NB | GWR-NB | Interpretation |
|--------|-----------|--------|----------------|
| AICc | higher | lower | GWR fits better |
| Deviance | higher | lower | Less unexplained variation |
| Pseudo-R² | lower | higher | More variance explained |

A pseudo-R² above 0.9 for the GWR model (versus, say, 0.18 for the global model)
indicates dramatic improvement from allowing spatially varying coefficients.

## Spatial autocorrelation in residuals

The Moran I test checks whether the residuals from the fitted model show a
spatial pattern. Significant positive autocorrelation in GWR residuals would
suggest the bandwidth is too large (over-smoothing).

```{r}
#| label: moran-residuals
#| eval: false

library(spdep)
library(gwzinbr)
data(southkorea_covid19)

# Assume mod is already fitted
resid_deviance <- mod$residuals[, 3]

# Build spatial weights (k-NN with k = 4)
coords <- cbind(southkorea_covid19$x, southkorea_covid19$y)
knn    <- knn2nb(knearneigh(coords, k = 4))
lw     <- nb2listw(knn, style = "W")

# Moran test (Monte Carlo)
moran.mc(resid_deviance, listw = lw, nsim = 999)
```

**Interpreting the result:**

| Moran I | p-value | Interpretation |
|---------|---------|----------------|
| $\approx 0$ | $> 0.05$ | Residuals are spatially random — GWR has removed spatial structure ✓ |
| $> 0$ | $< 0.05$ | Positive clustering remains — consider smaller $h$ or additional covariates |
| $< 0$ | $< 0.05$ | Negative autocorrelation — bandwidth may be too small |

## Non-stationarity test {#sec-nonstationarity}

The non-stationarity test assesses, for each predictor, whether the spatial
variation in local coefficients is statistically significant.

```{r}
#| label: nonstat-demo
#| eval: false

mod$non_stationary_test
#            variable   F_stat  df1    df2  p_value
# (Intercept)           2.341   4.2   91.8    0.032
# Morbidity             3.112   4.2   91.8    0.004
# high_sch_p            1.067   4.2   91.8    0.390
# ...
```

A covariate with a **non-significant** non-stationarity test could potentially be
modelled with a global (fixed) coefficient. Some authors use this to build
**mixed GWR models** where some coefficients are global and others are local —
though this requires a separate modelling step not currently in `gwzinbr`.

## Zero-inflation check {#sec-zero-check}

Before fitting a zero-inflated model, compare observed zeros to those expected
under the Negative Binomial distribution:

```{r}
#| label: zero-check
#| eval: false

library(MASS)
data(southkorea_covid19)
d <- southkorea_covid19

# Fit global NB
mod_nb_global <- glm.nb(
  n_covid1 ~ Morbidity + high_sch_p + Healthcare_access +
             diff_sd + Crowding + Migration + Health_behavior +
             offset(ln_total),
  data = d
)

mu_hat <- predict(mod_nb_global, type = "response")
theta  <- mod_nb_global$theta

zeros_obs <- sum(d$n_covid1 == 0)
zeros_exp <- sum((theta / (theta + mu_hat))^theta)

cat("Observed zeros:", zeros_obs, "\n")
cat("Expected zeros (NB):", round(zeros_exp, 1), "\n")
cat("Ratio:", round(zeros_obs / zeros_exp, 3), "\n")
```

**Decision rule:**

- Ratio close to 1 (observed ≈ expected): Negative Binomial is sufficient.
  Use `model = "negbin"`.
- Ratio substantially > 1: excess zeros present. Use `model = "zinb"` with
  `force = TRUE`.

::: {.callout-note}
`gwzinbr()` performs this check internally and prints a message. The
`force = TRUE` argument bypasses the automatic fallback.
:::

## Comparing global vs. GWR model

A complete comparison table can be assembled as follows:

```{r}
#| label: compare-models
#| eval: false

# Global NB (from MASS)
ll_global <- as.numeric(logLik(mod_nb_global))
ll0       <- as.numeric(logLik(update(mod_nb_global, . ~ 1)))
k_global  <- length(coef(mod_nb_global))
n         <- nobs(mod_nb_global)

aicc_global <- AIC(mod_nb_global) + 2 * k_global * (k_global + 1) /
               (n - k_global - 1)
r2_global   <- 1 - ll_global / ll0

# GWR-NB
gof <- mod$measures

data.frame(
  Model    = c("Global NB", "GWR-NB"),
  AICc     = round(c(aicc_global, gof["AICc"]), 2),
  Pseudo_R2 = round(c(r2_global,  gof["pct_ll"]), 3),
  Deviance = round(c(deviance(mod_nb_global), gof["deviance"]), 2)
)
```
