# Visualization {#sec-visualizacao}

`gwzinbr()` returns estimates as plain matrices. To produce maps, you need to
join these back to a spatial object (`sf`) that holds the geometry of each unit.
This chapter shows the complete workflow.

```{r}
#| label: setup-viz
#| cache: false
#| eval: false

library(gwzinbr)
library(sf)
library(ggplot2)
library(dplyr)
library(scales)
```

## Joining results to an sf object

```{r}
#| label: join-sf
#| eval: false

# Assume: dados_sf  = sf object (one row per district, same row order as model data)
#         mod       = fitted gwzinbr object
#         dados_modelo = plain data.frame used for fitting

# 1. Extract local coefficients
coefs <- as.data.frame(mod$gwr_param_estimates)

# 2. Extract t-test p-values
pvals <- as.data.frame(mod$t_test_gwr_param_estimates)

# 3. Extract residuals (deviance)
coefs$residuos <- mod$residuals[, 3]

# 4. Extract alpha (overdispersion)
coefs$alpha <- mod$alpha_estimates[, 1]

# 5. Join to sf
dados_sf <- bind_cols(dados_sf, coefs, pvals)
```

## Mapping local coefficients (masking non-significant)

The most informative GWR map shows **only statistically significant local estimates**,
masking non-significant districts in grey.

```{r}
#| label: mapa-coef
#| eval: false

# Helper: mask non-significant estimates
mask_insig <- function(estimate, pvalue, alpha = 0.10) {
  ifelse(pvalue < alpha, estimate, NA_real_)
}

dados_sf <- dados_sf %>%
  mutate(
    beta_x1_sig = mask_insig(x1, x1_pvalue, alpha = 0.10),
    beta_x2_sig = mask_insig(x2, x2_pvalue, alpha = 0.10)
  )

# Map for predictor x1
ggplot(dados_sf) +
  geom_sf(aes(fill = beta_x1_sig), color = "grey70", linewidth = 0.2) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "white",
    high     = "#b2182b",
    midpoint = 0,
    na.value = "grey85",
    name     = expression(hat(beta)[x1])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_line(color = "grey90")
  ) +
  labs(title = "Local coefficient — x1",
       subtitle = "Grey = not significant at 10%")
```

## Mapping residuals

```{r}
#| label: mapa-residuos
#| eval: false

ggplot(dados_sf) +
  geom_sf(aes(fill = residuos), color = "grey70", linewidth = 0.2) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "white",
    high     = "#b2182b",
    midpoint = 0,
    name     = "Deviance\nresidual"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Spatial distribution of deviance residuals — GWR-NB")
```

Residuals should appear spatially random after a well-fitted GWR model.
Clusters of large positive or negative residuals suggest missing covariates
or inappropriate bandwidth.

## Mapping the overdispersion parameter (alpha)

```{r}
#| label: mapa-alpha
#| eval: false

ggplot(dados_sf) +
  geom_sf(aes(fill = alpha), color = "grey70", linewidth = 0.2) +
  scale_fill_viridis_c(name = expression(hat(alpha)), option = "plasma") +
  theme_minimal(base_size = 13) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Local overdispersion parameter (α)",
       subtitle = "Higher α = more unexplained variability")
```

Regions with high $\hat{\alpha}$ are where the model fits poorly — the count
variance is much larger than predicted. These may warrant investigation for
missing spatial confounders.

## Multi-panel coefficient map

```{r}
#| label: mapa-multi
#| eval: false

library(patchwork)

make_coef_map <- function(df, col, pval_col, title, alpha = 0.10) {
  df <- df %>%
    mutate(coef_sig = mask_insig(.data[[col]], .data[[pval_col]], alpha))

  ggplot(df) +
    geom_sf(aes(fill = coef_sig), color = "grey70", linewidth = 0.15) +
    scale_fill_gradient2(
      low = "#2166ac", mid = "white", high = "#b2182b",
      midpoint = 0, na.value = "grey85", name = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text = element_blank(), axis.ticks = element_blank()) +
    labs(title = title)
}

p1 <- make_coef_map(dados_sf, "x1", "x1_pvalue", "x1")
p2 <- make_coef_map(dados_sf, "x2", "x2_pvalue", "x2")
p3 <- make_coef_map(dados_sf, "x3", "x3_pvalue", "x3")
p4 <- make_coef_map(dados_sf, "x4", "x4_pvalue", "x4")

(p1 | p2) / (p3 | p4)
```

## Descriptive summary of local estimates

```{r}
#| label: summary-coefs
#| eval: false

# Quick summary table (no mapping needed)
mod$descript_stats_gwr_param_estimates
#            Mean      SD     Min    Q25    Median   Q75    Max
# (Intercept) ...
# x1          ...
# x2          ...
```

This table is useful in publications when space is limited and you want to
communicate the range of local variation for each predictor.

## Convergence plot for bandwidth selection

After running `Golden()`, visualise the criterion as a function of bandwidth:

```{r}
#| label: gss-convergence-plot
#| eval: false

# Assume gss is the output of Golden()
iter <- gss$iterations

ggplot(iter, aes(x = h1, y = aic1)) +
  geom_point(color = "#2171b5", size = 2) +
  geom_point(aes(x = h2, y = aic2), color = "#b2182b", size = 2) +
  geom_vline(xintercept = gss$min_bandwidth, linetype = "dashed") +
  scale_x_continuous(name = "Bandwidth (h)") +
  scale_y_continuous(name = "AICc") +
  theme_minimal() +
  labs(title = "Golden Section Search convergence",
       subtitle = paste("Optimal h =", gss$min_bandwidth))
```
