# Case Study: COVID-19 in South Korea {#sec-estudo-caso}

This chapter presents a complete, reproducible analysis using the built-in
`southkorea_covid19` dataset. We follow the workflow from @sec-workflow:
check for overdispersion and excess zeros, fit a global NB as a reference,
assess spatial autocorrelation, select the bandwidth with `Golden()`, fit the
GWR-NB with `gwzinbr()`, and evaluate the results.

## The dataset

```{r}
#| label: dataset-overview
#| cache: false

library(gwzinbr)
library(MASS)
library(spdep)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)

data(southkorea_covid19)
d <- southkorea_covid19

cat("Observations:", nrow(d), "\n")
cat("Variables:", ncol(d), "\n")
cat("Column names:", paste(names(d), collapse = ", "), "\n")
```

```{r}
#| label: dataset-head
#| tbl-cap: "First rows of the southkorea_covid19 dataset."

head(d, 6) |>
  kable(digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

| Variable | Description |
|----------|-------------|
| `n_covid1` | Number of confirmed COVID-19 cases per district |
| `Morbidity` | Chronic disease morbidity rate |
| `high_sch_p` | Proportion of residents with high school education |
| `Healthcare_access` | Healthcare access index |
| `diff_sd` | Social distancing compliance score |
| `Crowding` | Population density / crowding index |
| `Migration` | Recent migration rate |
| `Health_behavior` | Composite health behaviour index |
| `x` | Longitude (decimal degrees) |
| `y` | Latitude (decimal degrees) |
| `ln_total` | Log of total population (offset) |

## Step 1: Check overdispersion and excess zeros

```{r}
#| label: step1-checks
#| tbl-cap: "Overdispersion check — Poisson residual deviance."

# Poisson model
mod_pois <- glm(
  n_covid1 ~ Morbidity + high_sch_p + Healthcare_access +
             diff_sd + Crowding + Migration + Health_behavior +
             offset(ln_total),
  data   = d,
  family = poisson()
)

disp_ratio <- deviance(mod_pois) / df.residual(mod_pois)

data.frame(
  Statistic = c("Residual deviance", "Degrees of freedom", "Dispersion ratio"),
  Value     = c(round(deviance(mod_pois), 1),
                df.residual(mod_pois),
                round(disp_ratio, 2))
) |> kable() |> kable_styling(full_width = FALSE)
```

::: {.callout-note}
A dispersion ratio much larger than 1 confirms overdispersion. The Negative
Binomial distribution is appropriate.
:::

```{r}
#| label: step1-zeros
#| tbl-cap: "Excess zeros check — observed vs expected under NB."

# Fit global NB
mod_nb_global <- glm.nb(
  n_covid1 ~ Morbidity + high_sch_p + Healthcare_access +
             diff_sd + Crowding + Migration + Health_behavior +
             offset(ln_total),
  data = d
)

mu_hat <- predict(mod_nb_global, type = "response")
theta  <- mod_nb_global$theta

zeros_obs <- sum(d$n_covid1 == 0)
zeros_exp <- round(sum((theta / (theta + mu_hat))^theta), 1)

data.frame(
  ` ` = c("Observed zeros", "Expected zeros (NB)", "Ratio"),
  Value = c(zeros_obs, zeros_exp, round(zeros_obs / zeros_exp, 2))
) |> kable(col.names = c("", "Value")) |>
     kable_styling(full_width = FALSE)
```

## Step 2: Global NB model (reference)

```{r}
#| label: step2-nb-coefs
#| tbl-cap: "Coefficients — Global Negative Binomial model."

broom::tidy(mod_nb_global) |>
  mutate(across(where(is.numeric), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2)) |>
  kable() |> kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
#| label: step2-nb-gof
#| tbl-cap: "Goodness-of-fit — Global NB."

ll   <- as.numeric(logLik(mod_nb_global))
ll0  <- as.numeric(logLik(update(mod_nb_global, . ~ 1)))
k    <- length(coef(mod_nb_global))
n    <- nobs(mod_nb_global)
aicc <- AIC(mod_nb_global) + 2 * k * (k + 1) / (n - k - 1)

data.frame(
  Statistic = c("Log-likelihood", "Pseudo-R² (McFadden)", "AIC", "AICc",
                "Theta (overdispersion)"),
  Value     = round(c(ll, 1 - ll / ll0, AIC(mod_nb_global), aicc,
                      mod_nb_global$theta), 3)
) |> kable() |> kable_styling(full_width = FALSE)
```

## Step 3: Spatial autocorrelation in residuals

```{r}
#| label: step3-moran
#| tbl-cap: "Moran's I test on global NB deviance residuals."
#| cache: true

coords <- cbind(d$x, d$y)
knn    <- knn2nb(knearneigh(coords, k = 4))
lw     <- nb2listw(knn, style = "W")

resid_nb <- residuals(mod_nb_global, type = "deviance")
moran_res <- moran.mc(resid_nb, listw = lw, nsim = 999)

data.frame(
  Statistic = c("Moran I", "p-value"),
  Value     = c(round(moran_res$statistic, 4), moran_res$p.value)
) |> kable() |> kable_styling(full_width = FALSE)
```

::: {.callout-important}
## Spatial autocorrelation detected
A significant Moran I confirms that the global NB model fails to capture the
spatial structure. A Geographically Weighted model is warranted.
:::

## Step 4: Bandwidth selection with `Golden()`

::: {.callout-note collapse="true"}
## Code: Golden Section Search (computationally intensive — ~10–30 min)

This code is shown for documentation purposes. The result (`h = 230`) is used
directly in the next step.

```{r}
#| label: step4-golden
#| eval: false

gss <- Golden(
  data       = d,
  formula    = n_covid1 ~ Morbidity + high_sch_p + Healthcare_access +
                          diff_sd + Crowding + Migration + Health_behavior,
  lat        = "y", long = "x",
  offset     = "ln_total",
  model      = "negbin",
  method     = "adaptive_bsq",
  bandwidth  = "aic",
  globalmin  = TRUE,
  distancekm = TRUE,
  force      = TRUE
)

gss$min_bandwidth  # >> 230
gss$iterations     # convergence table
```
:::

The search converged to an adaptive bandwidth of **h = 230 neighbours**, selected
by AICc. This means each local regression uses the 230 nearest districts.

## Step 5: Fit the GWR-NB model

::: {.callout-note collapse="true"}
## Code: GWR-NB fitting (computationally intensive — ~5–15 min)

```{r}
#| label: step5-fit
#| eval: false

mod_gwr <- gwzinbr(
  data       = d,
  formula    = n_covid1 ~ Morbidity + high_sch_p + Healthcare_access +
                          diff_sd + Crowding + Migration + Health_behavior,
  lat        = "y", long = "x",
  offset     = "ln_total",
  model      = "negbin",
  method     = "adaptive_bsq",
  h          = 230,
  distancekm = TRUE,
  force      = TRUE
)

saveRDS(mod_gwr, "_cache/mod_gwr_korea.rds")
```
:::

```{r}
#| label: step5-load
#| cache: false

# Load cached model (generated by script above)
mod_path <- "_cache/mod_gwr_korea.rds"
if (file.exists(mod_path)) {
  mod_gwr <- readRDS(mod_path)
} else {
  message("Run the fitting code above and save the model to _cache/mod_gwr_korea.rds")
}
```

## Step 6: Model comparison

```{r}
#| label: step6-compare
#| tbl-cap: "Comparison: Global NB vs GWR-NB."

if (exists("mod_gwr")) {
  gof <- mod_gwr$measures
  data.frame(
    Model      = c("Global NB", "GWR-NB"),
    AICc       = round(c(aicc, gof["AICc"]), 2),
    `Pseudo R2` = round(c(1 - ll / ll0, gof["pct_ll"]), 3),
    Deviance   = round(c(deviance(mod_nb_global), gof["deviance"]), 2)
  ) |> kable(col.names = c("Model", "AICc", "Pseudo R²", "Deviance")) |>
       kable_styling(bootstrap_options = "striped", full_width = FALSE)
}
```

## Step 7: Local coefficients — summary

```{r}
#| label: step7-coef-summary
#| tbl-cap: "Descriptive statistics of local GWR-NB coefficients."

if (exists("mod_gwr")) {
  as.data.frame(mod_gwr$descript_stats_gwr_param_estimates) |>
    round(4) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
}
```

```{r}
#| label: step7-nonstat
#| tbl-cap: "Non-stationarity test: is spatial variation significant?"

if (exists("mod_gwr")) {
  as.data.frame(mod_gwr$non_stationary_test) |>
    mutate(across(where(is.numeric), ~round(., 4))) |>
    kable() |>
    kable_styling(bootstrap_options = "striped", full_width = FALSE)
}
```

## Step 8: Spatial autocorrelation in GWR-NB residuals

```{r}
#| label: step8-moran-gwr
#| tbl-cap: "Moran's I — GWR-NB deviance residuals."
#| cache: true

if (exists("mod_gwr")) {
  resid_gwr <- mod_gwr$residuals[, 3]
  moran_gwr <- moran.mc(resid_gwr, listw = lw, nsim = 999)

  data.frame(
    Statistic = c("Moran I", "p-value"),
    Value     = c(round(moran_gwr$statistic, 4), moran_gwr$p.value)
  ) |> kable() |> kable_styling(full_width = FALSE)
}
```

::: {.callout-tip}
## Expected result
A non-significant Moran I ($p > 0.05$) for the GWR-NB residuals — versus the
significant autocorrelation found in the global NB residuals — confirms that the
GWR model successfully absorbed the spatial structure.
:::

## Step 9: Mapping local coefficients

The maps below require an `sf` object with the South Korea district geometries.
The code assumes you have loaded it as `korea_sf`, with rows in the same order
as `southkorea_covid19`.

```{r}
#| label: step9-maps
#| eval: false

library(sf)

# Add local estimates to the sf object
if (exists("mod_gwr") && exists("korea_sf")) {

  coefs <- as.data.frame(mod_gwr$gwr_param_estimates)
  pvals <- as.data.frame(mod_gwr$t_test_gwr_param_estimates)
  coefs$residuos <- mod_gwr$residuals[, 3]
  coefs$alpha    <- mod_gwr$alpha_estimates[, 1]

  korea_sf <- bind_cols(korea_sf, coefs, pvals)

  # Map: Crowding coefficient (significant only)
  korea_sf <- korea_sf |>
    mutate(Crowding_sig = ifelse(Crowding_pvalue < 0.10, Crowding, NA_real_))

  ggplot(korea_sf) +
    geom_sf(aes(fill = Crowding_sig), color = "grey70", linewidth = 0.2) +
    scale_fill_gradient2(
      low = "#2166ac", mid = "white", high = "#b2182b",
      midpoint = 0, na.value = "grey85",
      name = expression(hat(beta)[Crowding])
    ) +
    theme_minimal(base_size = 13) +
    theme(axis.text = element_blank(), axis.ticks = element_blank()) +
    labs(title = "Local coefficient — Crowding",
         subtitle = "Grey = not significant at 10%")
}
```
